Stellar Radiation is Critical for Regulating Star Formation and Driving Outflows in Low Mass Dwarf Galaxies

### 2019-03-18 thesis/chapters/ch_introduction.tex
New Figure for Leo P in introduction. cleaned up some bad figure labels and references. cleaned up some formatting

\newcommand{\code}{\textsc}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{./figures/intro/periodic_table}
  \caption{The periodic table of elements colored by the relative fraction of the nucleosynthetic processes that naturally occur in the Universe that produce each element. \textbf{Image Credit:} A modified version of the table created by Anna Frebel (\url{https://www.dropbox.com/sh/ib3agt2bkxqqk2u/AACbEw1warp74qzUh2AcWqfja?dl=0}), accessed on 03-18-19.}
\end{figure}

@@

As relevant to this work, we briefly summarize a few of the important nucleosynthetic channels shown in Figure~\ref{intro:fig:periodic table} and their sites. $\alpha$-elements, notably O, Mg, Si, Ca, and Ti, are produced during the lifetime of massive stars (O, Mg) or during the core collapse supernova explosion itself. In either case, these elements are synthesized through the capture of an $\alpha$ nucleus, progressing towards increasingly heavier nuclei in intervals of 4/2 in mass/atomic number (as $\alpha$ = $^{4}_2He$) from $^8_4Be$ to the stable $^{52}_{26}Fe$ and unstable $^{56}_{28}Ni$ (which days to $^{56}_{26}Fe$). The iron group elements, notably Fe, Mn, Co, and Ni, are produced predominately in Type Ia supernovae, with variations depending on a single vs. double degenerate scenario and the initial composition of exploding white dwarf(s) (WD).

@@

A full, detailed model of radiative cooling, heating, non-equillibrium chemistry, and dust is generally far too computationally expensive to include in galaxy-scale simulations. For this reason, many large-scale cosmological simulations assume that all gas is optically thin and that all molecules and ionization states exist in equillibrium. This allows for the quick use of tabulated look-up tables for radiative cooling and heating, often produced using \code{CLOUDY} \citep{Cloudy2013}, that depend on gas density, temperature, metallicity, and redshift (which sets the UVB heating rate). More complex simulations often track some form of $< 10$ species primordial chemistry in order to better account for non-equillibrium effects and $H_2$ production, though with noted increases in memory requirements and computational cost. Some of the most advance models to-date can account for well over 10$^2$ individual chemical reactions involving $>$ 20 elements and molecules \citep[e.g.][]{Richings2015,Richings2018}; but these are typically prohibitively expensive to run on massive scales. In this dissertation we adopt the chemistry and radiative cooling and heating physics models availble in the newly developed, open-source library \code{GRACKLE} \citep{GrackleMethod}. The exact physics included in our simulations are discussed in greater detail in Chapter~\ref{ch:chapter1}, but we note that improvements to \code{GRACKLE} have been made as the direct result of the research conducted during this dissertation.
%mention that I made improvements to this?

@@

In galaxy-scale hydrodynamics, star formation is typically treated as a process that occurs by a sub-grid model, with an assumed rate and assumed efficiency, in fluid elements that surpass a variety of threshold conditions. Usually, simulators set these thresholds to fully encompass gas whose Jeans-length (and thus dynamical behavior) becomes unresolved according to the widely-used \cite{Truelove1997} criterion. Often, this means a simple mass or number density threshold, on top of which simulators can also require that the local gas cells be in a converging flow ($\nabla \bm{v} < 0$), that gas surpass some virial threshold (describing how bound a cloud may be)
%cite?, 
or that gas meet a minimum $H_2$ mass fraction
%\citep[e.g.][]{}
. Once a fluid element is flagged as star forming, simulators have a few options to choose from in how to determine what fraction and with what rate that mass is converted into stars. Briefly, some form of the following equation is usually adopted,
\begin{equation}
\label{intro:eq:sfr}
 \rm{SFR} \propto 
\end{equation}
,. Some works adopt modifications of this simple perscription, with varying star formation efficiency ($\eta$) (refs), or XXXX. Others, for application in high-resolution simulations where the mass of stars that would form in a single timestep ($dt \times \rm{SFR}$) is far less than a single star particle mass, allow stars to form out of star forming fluid elements stochastically \citep[e.g.][]{Goldbaum2015}.

@@

In the simplest model, supernova feedback is treated as the injection of pure thermal energy (and mass) into the fluid element(s) that are host / closest to the chosen site of a supernova explosion. In self-consistent models of star formation and feedback these sites are usually actual star particles, but this does not have to be the case. The first models which included supernova feedback lacked the spatial / mass resolution to resolve individual (or even collective) SN events. This is the source of the commonly known ``overcooling" problem, whereby too little energy is injected into too much mass such that the affected region only reaches modest temperatures ($T \lesssim 10^5$~K, as opposed to $T > 10^7$~K, or so). This temperature is right around the peak of the cooling curve, and thus the energy is rapidly radiated away before the injected energy has any time to make a significant dynamical impact of the galaxy. This is still a problem today in large-box simulations of cosmological volumes which have insufficient resulution to resolve individual SNe. While some of the initial, ad-hoc fixes for this problem (e.g. turning off cooling for some time in SN affected fluid elements, decoupling SN affected SPH particles from the hydrodynamics for some time) are still in use today, some notable advancements have been made to employ some combination of kinetic, momentum, and thermal energy feedback to attempt to capture SNe in a more consistent fashion \citep[e.g.][]{Hopkins2014,Simpson2015,Hopkins2018}.

@@

Non-ionizing radiation also contributes to the regulation of star formation and the multi-phase ISM. Far ultraviolet (FUV) photons contribute to gas heating via the photoelectric heating of dust grains. This is a significant heating mechanism in higher metallicity environments, like the Milky Way \citep{Parravano2003,Wolfire2003}, but may still be important in setting temperature floor of cold dense gas in lower mass galaxies \citep{Forbes2016,Hu2017}. Lyman-Werner (LW) band radiation regulates the $H_2$ abundance, which is particularly important in low-metallicity environments where $H_2$ is a dominant coolant \citep[e.g.][]{Wolcott-Green2012}.\footnote{And is arguably even more important when using star formation perscriptions that depend upon the local $H_2$ mass fraction.} As the goal of this dissertation is to develop an as-complete-as-possible model of galactic chemical evolution, we account for each of the processes in some fashion, including the use of an adaptive ray-tracing radiative transfer scheme to track the stellar ionizing photons.

@@

Stellar winds, in addition to stellar radiation, are a potentially important source of pre-SN feedback that helps regulate star formation in / around the birth clouds of newly formed stars. However, the hot ($T>10^{6}$~K), fast ($v \sim 10^{3}$~km~s$^{-1}$) winds from massive stars \citep{Weaver1977} are challenging to model self-consistently in galaxy-scale simulations. For this reason, there has been comparatively little research on how they affect global galaxy properties, even though some models do include their energy injection in an integrated / approximate fashion \citep[e.g.][]{FIRE}. However, stellar winds, from both massive stars and weaker winds from the AGB phase of low mass stars, are important sources of metal enrichment. For these reasons we do include a simplified model for stellar wind feedback in this disseration.\footnote{Although we do not have conclusive results to this end, initial examination of our simulations has shown that our stellar wind model is only globally important in experimental simulations where radiation feedback is ignored. It is thus likely subdominant to radiation feedback.}

@@

While there is no real sharp transition where a galaxy is considered a ``dwarf" galaxy, historical definitions are based on luminosity, using the SMC and LMC as rough anchor points for the upper-end of the dwarf galaxy scale. \cite{Bullock2017} suggests a deliniation based on stellar mass. For this dissertation, we follow this definition and consider galaxies with $M_* < 10^9$~M$_{\odot}$ as dwarf galaxies, and galaxies with $M_* < 10^5$~M$_{\odot}$ as UFDs. In general, dwarf galaxies are extremely dark matter dominated, with mass to light ratios over 10$^2$, \citep{SimonGeha2007,Strigari2009,Wolf2010}, convert gas into stars at a lower efficiency than more massive galaxies (cite), and, for the lowest mass dwarf galaxies, exhibit bursty star formation histories driven by stochasticity in the process of both star formation and the effects of stellar feedback (cite).

@@

Stellar feedback drives outflows, galactic winds, from galaxies across all scales (references). Dwarf galaxies in particular, with their relatively shallow potential wells, are expected to drive effective galactic winds \citep{MacLowFerrara1999} with mass loading factors ($\eta$ = $\dot{\rm{M}}_{\rm out} / {\rm SFR}$) above XXX \citep{Muratov2015,Christensen2018}. Not only does this correspond to a significant amount of mass outflow from these galaxies, but these winds should also carry out a substantial amount of metals from these galaxies. \cite{Kirby2011-metals} used measurements of the stellar abundances in Local Group dSph's and their resolved star formation histories to determine that their stars contain less than 5\% of the metals synthesized in each galaxy. However, since these galaxies have lost their gas through environmental processes as satellites of the Milky Way (ram pressure and tidal stripping), it was unclear if this loss could be attributed entirely to feedback. However, using gas and stellar abundances \cite{McQuinn2015} showed that Leo P -- a gas-rich Local Group dwarf galaxy -- has also ejected $\sim$95\% of its metals. This is significant evidence for metal-enriched, feedback-driven galactic winds operating in these low mass dwarf galaxies.

@@

\subsection{Leo P: A Case Study}

\textbf{Include a figure? Maybe from recent spectra paper? Or HI from ALFALFA? or both?}

We dedicate this section to a specific Local Group dwarf galaxy, Leo P, as its observed properties motivate the initial conditions of the simulations developed in this dissertation. This galaxy is remarkable for being one of the lowest mass dwarf galaxies with observed, ongoing star formation. It is located at a distance of $1.62\pm 0.15$~Mpc \citep{McQuinn2015a}, and was first characterized as part of the ALFALFA survey \citep{Giovanelli2013,Rhode2013,Skillman2013,McQuinn2013,Bernstein-Cooper2014}. These studies have characterized its star formation history as low and roughly continuous over the age of the Universe (SFR = $4.3\times 10^{-5}$~M$_{\odot}$~yr$^{-1}$, \cite{McQuinn2015a}), its stellar mass ($M_{*} = 5.7 \times 10^{5}$~M$_{\odot}$, \cite{McQuinn2013}), its HI and dynamical mass ($M_{\rm HI}(r<r_{\rm HI}) = 9.5\times 10^{5}$~M$_{\odot}$, $M_{\rm dyn}(r<r_{\rm HI}) = 2.6 \times 10^{7}$~M$_{\odot}$, $r_{\rm HI} \sim 500$~pc, \cite{Bernstein-Cooper2014}), and gas-phase abundances from an HII region (12 + log(O/H) = 7.17$\pm$0.04 \cite{Skillman2013}). Finally, we chose to focus on a Leo P like galaxy in our simulations in part to allow for comparisons to the work in \cite{McQuinn2015}, which makes an accounting of the total metals (as traced by O) contained in the ISM and stars in Leo P, estimating the fraction that must have been ejected over the lifetime of the galaxy. This is an important measurement as prior estaimtes of this number for low mass dwarf galaxies were restricted to Milky Way dSph's, which are devoid of gas and contaminated by environmental affects. As of this writing, Leo P is one of the only low mass dwarf galaxies with both gas-phase and stellar metal abundance measurements, but there is ongoing work to greatly expand this sample in the near future.

@@

This dissertation utilizes high-resolution hydrodynamics simulations that are expensive to run in both walltime ($\sim$ weeks to months) and computational time ($\sim 10^{5-6}$ CPU hours). For this reason,

>>>>

\newcommand{\code}{\textsc}
\newcommand{\Hmolecular}{H$_2$}
%\newcommand{\hii}{H\sc{I}}
%\newcommand{\hii}{H{\sc{II}}}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{./figures/intro/periodic_table}
  \caption{The periodic table of elements colored by the relative fraction of the nucleosynthetic processes that naturally occur in the Universe that produce each element. \textbf{Image Credit:} A modified version of the table created by Anna Frebel (\url{https://www.dropbox.com/sh/ib3agt2bkxqqk2u/AACbEw1warp74qzUh2AcWqfja?dl=0}), accessed on 03-18-19.}
  \label{intro:fig:periodic table}
\end{figure}

@@

As relevant to this work, we briefly summarize a few of the important nucleosynthetic channels shown in Figure~\ref{intro:fig:periodic table} and their sites. $\alpha$-elements, notably O, Mg, Si, Ca, and Ti, are produced during the lifetime of massive stars (O, Mg) or during the core collapse supernova explosion itself. In either case, these elements are synthesized through the capture of an $\alpha$ nucleus, progressing towards increasingly heavier nuclei in intervals of 4/2 in mass/atomic number (as $\alpha$ = $^{4}_2$He) from $^8_4$Be to the stable $^{52}_{26}$Fe and unstable $^{56}_{28}$Ni (which days to $^{56}_{26}$Fe). The iron group elements, notably Fe, Mn, Co, and Ni, are produced predominately in Type Ia supernovae, with variations depending on a single vs. double degenerate scenario and the initial composition of exploding white dwarf(s) (WD).

@@

A full, detailed model of radiative cooling, heating, non-equillibrium chemistry, and dust is generally far too computationally expensive to include in galaxy-scale simulations. For this reason, many large-scale cosmological simulations assume that all gas is optically thin and that all molecules and ionization states exist in equillibrium. This allows for the quick use of tabulated look-up tables for radiative cooling and heating, often produced using \code{CLOUDY} \citep{Cloudy2013}, that depend on gas density, temperature, metallicity, and redshift (which sets the UVB heating rate). More complex simulations often track some form of $< 10$ species primordial chemistry in order to better account for non-equillibrium effects and \Hmolecular production, though with noted increases in memory requirements and computational cost. Some of the most advance models to-date can account for well over 10$^2$ individual chemical reactions involving $>$ 20 elements and molecules \citep[e.g.][]{Richings2015,Richings2018}; but these are typically prohibitively expensive to run on massive scales. In this dissertation we adopt the chemistry and radiative cooling and heating physics models availble in the newly developed, open-source library \code{GRACKLE} \citep{GrackleMethod}. The exact physics included in our simulations are discussed in greater detail in Chapter~\ref{ch:chapter1}, but we note that improvements to \code{GRACKLE} have been made as the direct result of the research conducted during this dissertation.
%mention that I made improvements to this?

@@

In galaxy-scale hydrodynamics, star formation is typically treated as a process that occurs by a sub-grid model, with an assumed rate and assumed efficiency, in fluid elements that surpass a variety of threshold conditions. Usually, simulators set these thresholds to fully encompass gas whose Jeans-length (and thus dynamical behavior) becomes unresolved according to the widely-used \cite{Truelove1997} criterion. Often, this means a simple mass or number density threshold, on top of which simulators can also require that the local gas cells be in a converging flow ($\nabla \bm{v} < 0$), that gas surpass some virial threshold (describing how bound a cloud may be)
%cite?, 
or that gas meet a minimum \Hmolecular mass fraction
%\citep[e.g.][]{}
. Once a fluid element is flagged as star forming, simulators have a few options to choose from in how to determine what fraction and with what rate that mass is converted into stars. Briefly, some form of the following equation is usually adopted,
\begin{equation}
\label{intro:eq:sfr}
 \rm{SFR} \propto 
\end{equation}
,. Some works adopt modifications of this simple perscription, with varying star formation efficiency ($\eta$) (refs), or XXXX. Others, for application in high-resolution simulations where the mass of stars that would form in a single timestep ($dt \times \rm{SFR}$) is far less than a single star particle mass, allow stars to form out of star forming fluid elements stochastically \citep[e.g.][]{Goldbaum2015}.

@@

In the simplest model, supernova feedback is treated as the injection of pure thermal energy (and mass) into the fluid element(s) that are host / closest to the chosen site of a supernova explosion. In self-consistent models of star formation and feedback these sites are usually actual star particles, but this does not have to be the case. The first models which included supernova feedback lacked the spatial / mass resolution to resolve individual (or even collective) SN events. This is the source of the commonly known ``overcooling" problem, whereby too little energy is injected into too much mass such that the affected region only reaches modest temperatures ($T \lesssim 10^5$~K, as opposed to $T > 10^7$~K, or so). This temperature is right around the peak of the cooling curve, and thus the energy is rapidly radiated away before the injected energy has any time to make a significant dynamical impact of the galaxy. This is still a problem today in large-box simulations of cosmological volumes which have insufficient resulution to resolve individual SNe. While some of the initial, ad-hoc fixes for this problem (e.g. turning off cooling for some time in SN affected fluid elements, decoupling SN affected SPH particles from the hydrodynamics for some time) are still in use today, some notable advancements have been made to employ some combination of kinetic, momentum, and thermal energy feedback to attempt to capture SNe in a more consistent fashion \citep[e.g.][]{FIRE,Simpson2015,Hopkins2018}.

@@

Non-ionizing radiation also contributes to the regulation of star formation and the multi-phase ISM. Far ultraviolet (FUV) photons contribute to gas heating via the photoelectric heating of dust grains. This is a significant heating mechanism in higher metallicity environments, like the Milky Way \citep{Parravano2003,Wolfire2003}, but may still be important in setting temperature floor of cold dense gas in lower mass galaxies \citep{Forbes2016,Hu2017}. Lyman-Werner (LW) band radiation regulates the \Hmolecular abundance, which is particularly important in low-metallicity environments where \Hmolecular is a dominant coolant \citep[e.g.][]{Glover2003,Wolcott-Green2012}.\footnote{And is arguably even more important when using star formation perscriptions that depend upon the local \Hmolecular mass fraction.} As the goal of this dissertation is to develop an as-complete-as-possible model of galactic chemical evolution, we account for each of the processes in some fashion, including the use of an adaptive ray-tracing radiative transfer scheme to track the stellar ionizing photons.

@@

Stellar winds, in addition to stellar radiation, are a potentially important source of pre-SN feedback that helps regulate star formation in / around the birth clouds of newly formed stars. However, the hot ($T>10^{6}$~K), fast ($v \sim 10^{3}$~km~s$^{-1}$) winds from massive stars \citep{Weaver1977} are challenging to model self-consistently in galaxy-scale simulations. For this reason, there has been comparatively little research on how they affect global galaxy properties, even though some models do include their energy injection in an integrated / approximate fashion \citep[e.g.][]{FIRE}. However, stellar winds, from both massive stars and weaker winds from the AGB phase of low mass stars, are important sources of metal enrichment. For these reasons we include stellar wind feedback in this disseration, albeit in a simplified fashion.\footnote{Although we do not have conclusive results to this end, initial examination of our simulations has shown that our stellar wind model is only globally important in experimental simulations where radiation feedback is ignored. It is thus likely subdominant to radiation feedback.}

@@

While there is no real sharp transition where a galaxy is considered a ``dwarf" galaxy, historical definitions are based on luminosity, using the SMC and LMC as rough anchor points for the upper-end of the dwarf galaxy scale. \cite{Bullock2017} suggests a deliniation based on stellar mass. For this dissertation, we follow this definition and consider galaxies with $M_* < 10^9$~M$_{\odot}$ as dwarf galaxies, and galaxies with $M_* < 10^5$~M$_{\odot}$ as UFDs. In general, dwarf galaxies are extremely dark matter dominated, with mass to light ratios over 10$^2$, \citep{SimonGeha2007,Strigari2008,Wolf2010}, convert gas into stars at a lower efficiency than more massive galaxies (cite), and, for the lowest mass dwarf galaxies, exhibit bursty star formation histories driven by stochasticity in the process of both star formation and the effects of stellar feedback (cite).

@@

Stellar feedback drives outflows, galactic winds, from galaxies across all scales (references). Dwarf galaxies in particular, with their relatively shallow potential wells, are expected to drive effective galactic winds \citep{MacLowFerrara1999} with mass loading factors ($\eta$ = $\dot{\rm{M}}_{\rm out} / {\rm SFR}$) above 100 \citep{Muratov2015,Christensen2018}. Not only does this correspond to a significant amount of mass outflow from these galaxies, but these winds should also carry out a substantial amount of metals from these galaxies. \cite{Kirby2011-metals} used measurements of the stellar abundances in Local Group dSph's and their resolved star formation histories to determine that their stars contain less than 5\% of the metals synthesized in each galaxy. However, since these galaxies have lost their gas through environmental processes as satellites of the Milky Way (ram pressure and tidal stripping), it was unclear if this loss could be attributed entirely to feedback. However, using gas and stellar abundances \cite{McQuinn2015} showed that Leo P -- a gas-rich Local Group dwarf galaxy -- has also ejected $\sim$95\% of its metals (see Section~\ref{intro:sec:Leo P}). This is significant evidence for metal-enriched, feedback-driven galactic winds operating in these low mass dwarf galaxies.

@@

\subsection{Leo P: A Case Study}
\label{intro:sec:Leo P}

\begin{figure}
 \centering
 \includegraphics[width=0.975\linewidth]{figures/intro/Leo_P}
 \caption{The Leo P dwarf galaxy. \textbf{Left:} \hi contours from the VLA at 32'' resolution (the lowest resolution available) overlaid on top of an optical LBT image. This image is adopted from the top-left panel of Figure 4 in \cite{Bernstein-Cooper2014}. \textbf{Right:} H$\alpha$ emission (contours) overlaid on top of an HST optical image. The \hii region analyzed in \cite{McQuinn2015} to obtain gas-phase metallicities is the blue clump of H$\alpha$ at the bottom of the image. This image is adopted from Figure 10 of \cite{Evans2019}. \textit{Note:} the veritcal axes are inverted between the left and right panels.}
 \label{intro:fig:Leo P}
\end{figure}

We dedicate this section to a specific Local Group dwarf galaxy, Leo P, as its observed properties motivate the initial conditions of the simulations developed in this dissertation. This galaxy is remarkable for being one of the lowest mass dwarf galaxies with observed, ongoing star formation. It is located at a distance of $1.62\pm 0.15$~Mpc \citep{McQuinn2015a}, and was first characterized as part of the ALFALFA survey \citep{Giovanelli2013,Rhode2013,Skillman2013,McQuinn2013,Bernstein-Cooper2014}. These studies have characterized its star formation history as low and roughly continuous over the age of the Universe (SFR = $4.3\times 10^{-5}$~M$_{\odot}$~yr$^{-1}$, \cite{McQuinn2015a}), its stellar mass ($M_{*} = 5.7 \times 10^{5}$~M$_{\odot}$, \cite{McQuinn2013}), its \hi and dynamical mass ($M_{\rm HI}(r<r_{\rm HI}) = 9.5\times 10^{5}$~M$_{\odot}$, $M_{\rm dyn}(r<r_{\rm HI}) = 2.6 \times 10^{7}$~M$_{\odot}$, $r_{\rm HI} \sim 500$~pc, \cite{Bernstein-Cooper2014}), and gas-phase abundances from an \hii region (12 + log(O/H) = 7.17$\pm$0.04 \cite{Skillman2013}). HI, optical, and H$\alpha$ observations of Leo P are given in Figure~\ref{intro:fig:Leo P}. Finally, we chose to focus on a Leo P like galaxy in our simulations in part to allow for comparisons to the work in \cite{McQuinn2015}, which makes an accounting of the total metals (as traced by O) contained in the ISM and stars in Leo P, estimating the fraction that must have been ejected over the lifetime of the galaxy. This is an important measurement as prior estaimtes of this number for low mass dwarf galaxies were restricted to Milky Way dSph's, which are devoid of gas and contaminated by environmental affects. As of this writing, Leo P is one of the only low mass dwarf galaxies with both gas-phase and stellar metal abundance measurements, but there is ongoing work to greatly expand this sample in the near future.

@@

This dissertation utilizes high-resolution hydrodynamics simulations that are expensive to run in both walltime ($\sim$ weeks to months) and computational time ($\sim 10^{5-6}$ CPU hours). For this reason, it is computationally infeasible to directly explore many of the uncertainties associated with each of the components of the model as discussed in Section~\ref{intro:sec:ingredients}. For this reason, substantially cheaper semi-analytic models are valuable tools that can be used to better understand galactic evolution.
